MADA Final Project

### 2023-02-24 products/manuscript/Manuscript.qmd
Submit Part 2

---
title: "MADA Final Project"
subtitle: "Trends in PrEP Use in the US, 2012-2022"
author: Nicole Luisi
date: "`r Sys.Date()`"
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../dataanalysis_template_references.bib
csl: ../apa.csl
---

# PART 1

**Overview**:

The purpose of this analysis is to measure the public health impact of pre-exposure prophylaxis (PrEP) in the US over the past 10 years. Previous research by [Sullivan et al. (2018)](https://doi.org/10.1016/j.annepidem.2018.06.009) evaluated trends in the prevalence of PrEP use from 2012-2017. Outcomes include period prevalence of PrEP use and PrEP-to-need ratio (PnR) by year. This analysis will extend the previous work to include data through the year 2022; subgroups will include region, state, age group, gender, and race/ethnicity *(race/ethnicity PrEP data were not available at the time of the previous analysis and were not included)*. [@SULLIVAN2018833]

*Note: New diagnoses data are currently available through 2020; PrEP data through 2021. If 2021-2022 data are not available by the time the analysis needs to be completed, the analysis will be temporarily restricted to 2012-2020.*

+ HIV New Diagnoses: [AIDSVu](https://aidsvu.org/) [@Sullivan2020] 
+ PrEP and PnR: IQVIA (Quintiles IMS Holdings)
+ Annual Population Estimates: National Historical Geographic Information System (NHGIS), American Community Surveys (ACS), US Census Bureau

+ N/% of annual PrEP users by age, gender, race/ethnicity, and region
+ Annual prevalence of PrEP use per 100,000 population by age, gender, race/ethnicity, region, and state
+ Estimated annual percent change (EAPC) for prevalence
  - Calculated using [Joinpoint Regression Software (R-Callable Command-Line Version 4.9.0.0)](https://surveillance.cancer.gov/help/joinpoint/tech-help/citation)
+ Prep-to-need ratio by age, gender, race/ethnicity, and region
+ Other modeling TBD based on course modules

The structure below is one possible setup for a data analysis project (including the course project). For a manuscript, adjust as needed. 
You don't need to have exactly these sections, but the content covering those sections should be addressed.

This uses MS Word as output format. [See here](https://quarto.org/docs/output-formats/ms-word.html) for more information. You can switch to other formats, like html or pdf. See [the Quarto documentation](https://quarto.org/) for other formats.

```{r, echo=FALSE, message=FALSE}
# load a few R packages
#library(here)
#library(knitr)
```

# Summary/Abstract
_Write a summary of your project._

@@

## Description of data and data source
_Describe what the data is, what it contains, where it is from, etc. Eventually this might be part of a methods section._

@@

To cite other work (important everywhere, but likely happens first in introduction), make sure your references are in the bibtex file specified in the YAML header above (here `dataanalysis_template_references.bib`) and have the right bibtex key. Then you can include like this:

Examples of reproducible research projects can for instance be found in []

@@

## Data aquisition
_As applicable, explain where and how you got the data. If you directly import the data from an online source, you can combine this section with the next._

## Data import and cleaning
_Write code that reads in the file and cleans it so it's ready for analysis. Since this will be fairly long code for most datasets, it might be a good idea to have it in one or several R scripts. If that is the case, explain here briefly what kind of cleaning/processing you do, and provide more details and well documented code somewhere (e.g. as supplement in a paper). All materials, including files that contain code, should be commented well so everyone can follow along._

@@

_Use a combination of text/tables/figures to explore and describe your data. Show the most important descriptive results here. Additional ones should go in the supplement. Even more can be in the R and Quarto files that are part of your project._

@tbl-summarytable shows a summary of the data.

Note the loading of the data providing a **relative** path using the `../../` notation. (Two dots means a folder up). You never want to specify an **absolute** path like `C:\ahandel\myproject\results\` because if you share this with someone, it won't work for them since they don't have that path. You can also use the `here` R package to create paths. See examples of that below.

```{r}
#| label: tbl-summarytable
#| tbl-cap: "Data summary table."
#| echo: FALSE
#resulttable=readRDS("../../results/summarytable.rds")
#knitr::kable(resulttable)
```

>>>>

---
title: "MADA Final Project"
subtitle: "Trends in New HIV Diagnoses in the US, 2008-2022"
author: Nicole Luisi
date: "`r Sys.Date()`"
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../dataanalysis_template_references.bib
csl: ../apa.csl
---

Note: There seem to be  lot of issues getting tables to render well in Word doc format, need to come back to this. I tried 5 or so different methods and they all look terrible when in a doc vs html. Also having issues with resolving crossrefs.

```{r, echo=FALSE, message=FALSE}
# load R packages
library(here) 
library(ggplot2)
library(skimr) 
library(plotly)
library(gt)
library(knitr)
```

# Summary/Abstract

The purpose of this analysis is to analyze US trends in new HIV diagnoses over the past 10 years. Previous research by [Chapin-Bardales et al. (2017)](https://doi.org/10.1016/j.annepidem.2017.04.002) evaluated trends in AIDS diagnosis rates by race/ethnicity between 1984-2013. This analysis will focus on new HIV diagnosesbetween  2008-2022; subgroups will include region, age group, gender, and race/ethnicity. [@Chapin2017]

@@

## Description of data and data source

*Note: New diagnoses data are currently available through 2020. If 2021-2022 data are not available by the time the analysis needs to be completed, the analysis will be temporarily restricted to 2012-2020.*

+ HIV New Diagnoses: [AIDSVu](https://aidsvu.org/) [@Sullivan2020]

+ Rate (per 100,000) of new HIV diagnoses by age, gender, and race/ethnicity (national, regional)
+ Rate ratios for new HIV diagnoses by age, gender, and race/ethnicity (national, regional)
+ Estimated annual percent change (EAPC) for incidence
  - Calculated using [Joinpoint Regression Software (R-Callable Command-Line Version 4.9.0.0)](https://surveillance.cancer.gov/help/joinpoint/tech-help/citation)
+ Other modeling TBD based on course modules

@@

@@

## Data aquisition
Data files were downloaded from [AIDSVu](https://aidsvu.org/) and placed in the `data/raw_data/AIDSVu` subfolder. The new diagnosis files consist of one xlsx file per year for each level (national, regional, etc.).

## Data import and cleaning
The `processingcode.R` file includes the following steps:
+ Read in xlsx data files and combine all years for each level
+ Subset variables of interest and rename variables so all levels are consistent 
+ View/explore data
+ Calculate new rate ratio variables for subgroups 
+ Prepare combined national and regional file with all variables

@@

@tbl-natsummarytable shows a full summary of the National data.

```{r}
#| label: tbl-natsummarytable
#| tbl-cap: "National data summary table."
#| echo: FALSE
alldata<-readRDS(here("data", "processed_data", "allnewdx.rds"))
national<-alldata %>% filter(Geo %in% c("United States"))
national %>% gt::gt() %>% 
  gt::tab_header(title = "National Data") %>%
  gt::tab_style(style = list(gt::cell_fill(color = "#7cb9e8"),
                             gt::cell_text(weight = "bold")),
                locations = gt::cells_body(columns = Year))
```
@fig-natplots shows exploratory plots for the national rates by subgroup.

```{r}
#| label: fig-natplots
#| fig-cap: "National rates by subgroup"
#| echo: FALSE
national<-alldata %>% filter(Geo %in% c("United States"))
natfig <- plot_ly(national, x = ~Year, y = ~Overall_Rate, name = "Overall", type="scatter", mode = "lines", line = list(width = 3, dash = "dash", color = "gray")) 
natfig <- natfig %>% add_trace(y = ~Male_Rate, name = "Male", mode = 'lines', line = list(width = 3, dash = "solid", color = "blue"))
natfig <- natfig %>% add_trace(y = ~Female_Rate, name = "Female", mode = 'lines', line = list(width = 3, dash = "solid", color = "green"))
natfig %>% layout(title="National New Diagnosis Rate by Sex", legend=list(title=list(text='Sex')), 
                  xaxis = list(title = 'Year'),
                  yaxis = list(title = 'Rate per 100,000'))

natfig <- plot_ly(national, x = ~Year, y = ~Overall_Rate, name = "Overall", type="scatter", mode = "lines", line = list(width = 3, dash = "dash", color = "gray")) 
natfig <- natfig %>% add_trace(y = ~Black_Rate, name = "Black", mode = 'lines', line = list(width = 3, dash = "solid", color = "blue"))
natfig <- natfig %>% add_trace(y = ~White_Rate, name = "White", mode = 'lines', line = list(width = 3, dash = "solid", color = "darkgreen"))
natfig <- natfig %>% add_trace(y = ~Hispanic_Rate, name = "Hispanic", mode = 'lines', line = list(width = 3, dash = "solid", color = "darkred"))
natfig <- natfig %>% add_trace(y = ~Asian_Rate, name = "Asian", mode = 'lines', line = list(width = 3, dash = "solid", color = "cornflowerblue"))
natfig <- natfig %>% add_trace(y = ~AIAN_Rate, name = "AIAN", mode = 'lines', line = list(width = 3, dash = "solid", color = "brown1"))
natfig <- natfig %>% add_trace(y = ~MultRace_Rate, name = "Multiracial", mode = 'lines', line = list(width = 3, dash = "solid", color = "purple"))
natfig <- natfig %>% add_trace(y = ~NHPI_Rate, name = "NHPI", mode = 'lines', line = list(width = 3, dash = "solid", color = "magenta"))
natfig %>% layout(title="National New Diagnosis Rate by Race/Ethnicity", legend=list(title=list(text='Race/Ethnicity')), 
                  xaxis = list(title = 'Year'),
                  yaxis = list(title = 'Rate per 100,000'))

natfig <- plot_ly(national, x = ~Year, y = ~Overall_Rate, name = "Overall", type="scatter", mode = "lines", line = list(width = 3, dash = "dash", color = "gray")) 
natfig <- natfig %>% add_trace(y = ~Age13t24_Rate, name = "Age 13-24", mode = 'lines', line = list(width = 3, dash = "solid", color = "blue"))
natfig <- natfig %>% add_trace(y = ~Age25t34_Rate, name = "Age 25-34", mode = 'lines', line = list(width = 3, dash = "solid", color = "darkgreen"))
natfig <- natfig %>% add_trace(y = ~Age35t44_Rate, name = "Age 35-44e", mode = 'lines', line = list(width = 3, dash = "solid", color = "darkred"))
natfig <- natfig %>% add_trace(y = ~Age45t54_Rate, name = "Age 45-54", mode = 'lines', line = list(width = 3, dash = "solid", color = "cornflowerblue"))
natfig <- natfig %>% add_trace(y = ~Age55p_Rate, name = "Age 55+", mode = 'lines', line = list(width = 3, dash = "solid", color = "orange"))
natfig %>% layout(title="National New Diagnosis Rate by Age", legend=list(title=list(text='Age')), 
                  xaxis = list(title = 'Year'),
                  yaxis = list(title = 'Rate per 100,000'))
```

@tbl-natratios shows a summary of selected rate ratios for race/ethnicity and sex.

```{r}
#| label: tbl-natratios
#| tbl-cap: "National rate ratios by select subgroups"
#| echo: FALSE
national<-alldata %>% filter(Geo %in% c("United States"))
national_sub <- national %>% dplyr::select("Year", "Black_Rate", "White_Rate", "Hispanic_Rate", "black_white_rateratio", "hispanic_white_rateratio", "Male_Rate", "Female_Rate", "female_male_rateratio")
national_sub %>% gt::gt() %>% 
  gt::tab_header(title = "Rate Ratios") %>%
  gt::tab_style(style = list(gt::cell_fill(color = "#7cb9e8"),
                             gt::cell_text(weight = "bold")),
                locations = gt::cells_body(columns = Year))%>%
  gt::tab_style(style = list(gt::cell_fill(color = "#F1EB9C"),
                             gt::cell_text(weight = "bold")), 
                locations = gt::cells_body(columns = c(black_white_rateratio,hispanic_white_rateratio,female_male_rateratio)))
```

@tbl-regsummarytable shows a full summary of the Regional data.

```{r}
#| label: tbl-regsummarytable
#| tbl-cap: "Regional data summary table."
#| echo: FALSE
alldata<-readRDS(here("data", "processed_data", "allnewdx.rds"))
regional<-alldata %>% filter(Geo %in% c("Northeast", "Midwest", "South", "West"))
regional %>% gt::gt() %>% 
  gt::tab_header(title = "Regional Data") %>%
  gt::tab_style(style = list(gt::cell_fill(color = "#7cb9e8"),
                             gt::cell_text(weight = "bold")),
                locations = gt::cells_body(columns = Year))
```

@fig-regplots shows an exploratory plot for regional overall new diagnosis rates with female-male rate ratios.

```{r}
#| label: fig-regplots
#| tbl-cap: "Regional plots"
#| echo: FALSE
alldata<-readRDS(here("data", "processed_data", "allnewdx.rds"))
regional<-alldata %>% filter(Geo %in% c("Northeast", "Midwest", "South", "West"))
ggplot(regional, aes(x=Year, y=log(Overall_Rate), size=log(female_male_rateratio), color=Geo)) +
    geom_point(alpha=0.7) + scale_x_continuous(breaks=c(2008,2009,2010,2011,2012,2013,2014,2015,2016,2017,2018,2019,2020)) +
  ggtitle("Regional New Diagnosis Rates with Female-Male Rate Ratios") +
  theme(plot.title = element_text(hjust = 0.5)) +
  xlab("Year") + ylab("Log of Overall NewDx Rate")
```
